% !TeX root = ../main.tex

% 中英文摘要和关键字

\begin{abstract}

图神经网络通过沿图结构聚合多跳邻域信息，在节点分类、链接预测等任务中表现出较好的建模能力，已成为图数据学习中的重要方法。随着相关模型逐步走向推荐系统、实时风控等对响应时延较为敏感的应用场景，推理阶段的效率问题开始变得突出。多层消息传递会带来感受野的持续扩张，邻居检索与特征聚合开销随之迅速上升，模型因而难以直接用于大规模在线服务。另一方面，真实图中普遍存在噪声边、缺失边等结构缺陷，这些问题会在固定邻接矩阵的传播过程中累积到节点表示中，进一步影响下游任务的稳定性。

稀疏分解图神经网络（Sparse Decomposition of Graph Neural Networks，SDGNN）将教师模型的表示逼近写成单次稀疏矩阵乘法，在明显降低推理开销的同时保持了较好的近似能力。不过，这一方法对所有节点使用同一个正则系数来控制稀疏强度，没有反映节点之间实际存在的拓扑差异。对于连接范围较广、承担更多结构信息的枢纽节点，过强的稀疏约束容易删去关键邻居；对于局部连接较少的外围节点，过弱的稀疏约束又会留下较多冗余连接。统一稀疏控制在结构差异较强的图上往往难以同时兼顾精度与压缩效果，这也是 SDGNN 进一步改进时最需要处理的问题。

围绕这一局限，本文在 SDGNN 基础上提出多尺度自适应稀疏图神经网络（MSAS-GNN），根据节点所处的局部结构环境分配差异化稀疏预算。本文从图结构中提取谱间隙、节点谱能量、归一化局部图熵和节点中心性四类可计算指标，用于描述图的全局连通性、节点谱域参与强度和局部邻域结构特征，并据此构建频率、节点和跳距三个维度的自适应稀疏机制。频率维度采用节点谱能量的等权实例化作为统一频率接口，以避免引入与具体任务强绑定的额外统计量；节点维度结合节点谱能量与邻域结构复杂度构造节点级正则系数，使信息贡献更高的节点受到较弱稀疏约束，从而保留更多关键邻居；跳距维度结合多跳传播中信息贡献随距离递减的现象，对不同环层分配预算，并在均匀分配与几何衰减分配之间进行自适应选择。完成参数设计后，推理阶段仍保持单次稀疏矩阵乘法形式，主项复杂度维持在线性量级。

本文在 Cora、Citeseer、PubMed、ogbn-arxiv、Chameleon 和 Squirrel 六个静态节点分类基准数据集上对 MSAS-GNN 进行了验证。主实验与显著性检验均在 10 个随机种子下重复，并采用 Wilcoxon 符号秩检验评估差异。结果表明，MSAS-GNN 在六个数据集上均优于基线 SDGNN，且逐数据集差异均达到显著水平。在引文网络数据集上，Cora 的准确率达到 88.0\%，Citeseer 与 PubMed 相比 SDGNN 也均有提升，三个引文网络数据集平均提升约 1.3 个百分点；在大规模图 ogbn-arxiv 上，准确率由 74.27\% 提升至 75.13\%；在 Chameleon 和 Squirrel 上，准确率分别达到 65.6\% 和 55.9\%。为进一步检验本文方法在近年高效图学习方法下的相对表现，本文进一步补充 GraphSAINT、NodeFormer、DIFFormer、SGFormer 和 NAGphormer 等基线。补充结果显示，SGFormer 在多项数据集上是最接近本文方法的近期基线，在 ogbn-arxiv 上达到 74.96\%，在 Chameleon 和 Squirrel 上分别达到 65.2\% 和 55.4\%；MSAS-GNN 对应结果为 75.13\%、65.6\% 和 55.9\%，仍保持更高准确率。效率测试表明，MSAS-GNN 在 ogbn-arxiv 上相对标准消息传递基线实现约 13.6 倍加速，四个代表性数据集上的平均加速比约为 8.0 倍，峰值前向显存占用减少约 74.8\%。消融结果进一步说明，性能改善并不能由单一结构因子单独解释，节点级多因子正则与近邻优先的分层跳距预算需要共同作用，才能稳定体现出方法优势。

本文从图结构统计量出发，构建面向节点差异的稀疏分配方式，降低稀疏推理对全局统一预算的依赖。在保持线性推理复杂度的前提下，该方法提升了不同拓扑角色节点的适配性，也为大规模图上的高效推理提供了较为直接的实现路径。

  % 关键词用“英文逗号”分隔，输出时会自动处理为正确的分隔符
  \bnusetup{
    keywords = {图神经网络, 自适应稀疏化, 谱图分析, 节点异质性, 跳距预算分配},
  }
\end{abstract}

\begin{abstract*}

Graph Neural Networks (GNNs) have become an important tool for learning from graph-structured data because they build node representations by aggregating information from multi-hop neighborhoods. As these models move from benchmark studies to latency-sensitive applications such as recommender systems and real-time risk control, inference efficiency becomes a practical concern. Multi-layer message passing continuously enlarges the receptive field, which quickly increases the cost of neighbor retrieval and feature aggregation during inference and makes direct deployment difficult in large-scale online settings. In addition, real-world graphs often contain noisy edges and missing connections, and such structural defects can propagate through a fixed adjacency matrix and eventually affect the stability of downstream predictions.

Sparse Decomposition of Graph Neural Networks (SDGNN) rewrites the approximation of teacher representations as a single sparse matrix multiplication, substantially reducing inference cost while preserving good approximation quality. However, SDGNN uses one global regularization coefficient to control sparsity for all nodes, which does not reflect the topological differences among them. For hub nodes that carry richer structural information, overly strong sparsification may remove critical neighbors; for peripheral nodes with limited local connectivity, overly weak sparsification may keep redundant connections and reduce compression gains. This mismatch makes it difficult for a uniform sparsity policy to balance accuracy and efficiency on graphs with pronounced structural heterogeneity.

Starting from this limitation, this thesis develops the Multi-Scale Adaptive Sparse Graph Neural Network (MSAS-GNN) on top of SDGNN. Instead of assuming that all nodes should receive the same sparsity pressure, the proposed method allocates different sparsity budgets according to the local structural environment of each node. To support this design, four computable graph indicators are first extracted from the graph structure, namely spectral gap, node spectral energy, normalized local graph entropy, and node centrality. These quantities are then used to characterize global connectivity, spectral participation at the node level, and local neighborhood structure. Based on them, an adaptive sparsification scheme is built from three perspectives: frequency, node, and hop distance. In the frequency dimension, an equal-weight instantiation of node spectral energy is adopted as the common frequency interface so that the method does not rely on additional task-specific statistics. In the node dimension, node spectral energy and neighborhood structural complexity are combined to construct node-wise regularization coefficients, allowing nodes with higher information contribution to receive weaker sparsity constraints and retain more critical neighbors. In the hop-distance dimension, budgets are assigned across different neighborhood rings according to the observed decay of information contribution with hop distance, with adaptive selection between uniform allocation and geometric-decay allocation. After these parameters are determined, inference still takes the form of a single sparse matrix multiplication, so the dominant complexity remains at a linear scale.

Experiments are conducted on six static node classification benchmarks: Cora, Citeseer, PubMed, ogbn-arxiv, Chameleon, and Squirrel. All main experiments and significance tests are repeated with 10 random seeds, and the Wilcoxon signed-rank test is used to evaluate the differences. The results show that MSAS-GNN outperforms the baseline SDGNN on all six datasets, and every per-dataset improvement is statistically significant. On the citation-network benchmarks, the accuracy on Cora reaches 88.0\%, while Citeseer and PubMed also improve over SDGNN, giving an average gain of about 1.3 percentage points across the three datasets. On the large-scale graph ogbn-arxiv, the accuracy increases from 74.27\% to 75.13\%. On Chameleon and Squirrel, the accuracies reach 65.6\% and 55.9\%, respectively. To further evaluate the proposed method against recent efficient graph learning and structure-adaptive approaches, additional baselines including GraphSAINT, NodeFormer, DIFFormer, SGFormer, and NAGphormer are included. Among these recent baselines, SGFormer is the closest competitor in several settings, reaching 74.96\% on ogbn-arxiv and 65.2\% and 55.4\% on Chameleon and Squirrel, while MSAS-GNN achieves 75.13\%, 65.6\%, and 55.9\% on the same datasets. Efficiency evaluation shows that MSAS-GNN achieves about 13.6$\times$ speedup over the standard message-passing baseline on ogbn-arxiv, with an average speedup of about 8.0$\times$ across four representative datasets, while reducing peak forward GPU memory usage by about 74.8\%. Ablation results further indicate that the gain cannot be attributed to any single structural factor alone; instead, it emerges from the combined effect of multi-factor node-wise regularization and near-neighbor-priority layer-wise hop-budget allocation.

In summary, this thesis develops a node-differentiated sparsity allocation scheme from computable graph statistics, so that sparse inference no longer relies on a globally uniform budget. While preserving linear inference complexity, the proposed method adapts better to nodes with different topological roles and provides a practical route toward efficient inference on large-scale graphs.

  % Use comma as separator when inputting
  % 英文关键词首字母请大写
  \bnusetup{
    keywords* = {Graph Neural Network, Adaptive Sparsification, Spectral Graph Analysis, Node Heterogeneity, Hop-Distance Budget Allocation},
  }
\end{abstract*}